Seminar 5: Fitting and interpreting linear models (low volume)
==============================================================
### Lauren Chong

Load the data:
```{r}
library(lattice)
library(reshape2)
prDat <- read.table("GSE4051_data.tsv")
str(prDat, max.level = 0)
prDes <- readRDS("GSE4051_design.rds")
str(prDes)
```

Write a function to extract info for a list of genes:
```{r}
prepareData <- function(myGenes) {
  genesDat <- prDat[myGenes,]
  genesDat <- data.frame(
    gExp = as.vector(t(as.matrix(genesDat))),
    gene = factor(rep(myGenes, each = ncol(genesDat)), levels = myGenes)
  )
  genesDat <- suppressWarnings(data.frame(prDes, genesDat))
  return(genesDat)
}

# Test it out
(luckyGenes <- c("1419655_at","1438815_at"))
lDat <- prepareData(luckyGenes)
str(lDat)
head(lDat)

stripplot(gExp ~ devStage | gene, lDat,
          group = gType, jitter.data = TRUE,
          auto.key = TRUE, type = c('p', 'a'), grid = TRUE)
```

Borrow the makestripplot function from Jenny's code:
```{r}
makeStripplot <- function(myData, ...) {
  stripplot(gExp ~ devStage | gene, myData,
            group = gType, jitter.data = TRUE,
            auto.key = TRUE, type = c('p', 'a'), grid = TRUE, ...)
}
```

Further work
------------
Retry some similar analyses from the tutorial, but treat developmental stage as a numerical variable. Drop the last time point too, as per Jenny's observations:
```{r}
library(car)
prDes$age <- + recode(prDes$devStage, "'E16'=-2; 'P2'=2; 'P6'=6; 'P10'=10; '4_weeks'=28", as.factor.result = FALSE)
prDes <- prDes[prDes$devStage != "4_weeks",]
prDat <- prDat[,prDes$sidChar]
```

### Gene 1
Experiment with a single gene of interest:
```{r}
oDat <- prepareData("1448690_at")
xyplot(gExp ~ age | gene, oDat,
          group = gType, jitter.data = TRUE,
          auto.key = TRUE, type = c('p', 'a'), grid = TRUE)
fit.lm <- lm(gExp ~ age * gType, oDat)
summary(fit.lm)
```

Our fitted model tells us there is no significant interaction effect. Try a new model with no interaction, and compare the results using ANOVA:
```{r}
fit.lm.noint <- lm(gExp ~ age + gType, oDat)
summary(fit.lm.noint)
anova(fit.lm.noint, fit.lm)
```

The large p-value indicates that adding the interactions is unnecessary.

### Gene 2
Now try analysis with a more interesting gene:
```{r}
pDat <- prepareData("1429225_at")
xyplot(gExp ~ age | gene, pDat,
          group = gType, jitter.data = TRUE,
          auto.key = TRUE, type = c('p', 'a'), grid = TRUE)
```

This looks like it may be non-linear, so try comparing linear and quadratic age models:
```{r}
quad.fit.small <- lm(formula = gExp ~ age, data = pDat)
summary(quad.fit.small)
quad.fit <- lm(formula = gExp ~ age + I(age^2), data = pDat)
summary(quad.fit)
anova(quad.fit.small, quad.fit)
```

The p-value is very significant, meaning that the quadratic model is a big improvement. Now see if adding genotype changes anything:
```{r}
quad.fit.gtype <- lm(formula = gExp ~ gType + (age + I(age^2)), data = pDat)
summary(quad.fit.gtype)
anova(quad.fit, quad.fit.gtype)
```

So it looks like adding the genotype information is moderately informative. What about an interaction effect?
```{r}
quad.fit.int <- lm(formula = gExp ~ gType * (age + I(age^2)), data = pDat)
summary(quad.fit.int)
anova(quad.fit, quad.fit.int)
```

Adding an interaction between age and genotype appears to be a significant improvement. This is the best model we have tried for this gene.

### Conclusions
Using a numerical covariate for age allowed us to fit quadratic models which were more appropriate for the 2nd gene.