STAT 540 -- Seminar 7
========================================================

Looking at read count data from RNA-seq of 21 mice (2 strains).

Load packages:
```{r}
library(edgeR)
library(DESeq)
library(limma)
```

Load the data:
```{r}
dat <- read.table("bottomly_count_table.tsv", header = TRUE, row.names = 1)
des <- read.table("bottomly_phenodata.tsv", header = TRUE, row.names = 1)
dim(dat)
head(dat)
show(des)
```

edgeR analysis
--------------
First we will run our DE analysis using edgeR to model the expression.  

Create a DGEList separating the samples by strain, and then a design matrix for model fitting:
```{r}
group <- factor(as.numeric(des$strain))
dge.glm <- DGEList(counts = dat, group = group)
show(dge.glm$samples)
design <- model.matrix(~group)
show(design)
```

Estimate dispersion for each probe:
```{r}
dge.glm.com.disp <- estimateGLMCommonDisp(dge.glm, design, verbose = TRUE)
dge.glm.trend.disp <- estimateGLMTrendedDisp(dge.glm.com.disp, design)
dge.glm.tag.disp <- estimateGLMTagwiseDisp(dge.glm.trend.disp, design)
plotBCV(dge.glm.tag.disp)
```

Fit an adjusted linear model, run likelihood test:
```{r}
fit <- glmFit(dge.glm.tag.disp, design)
lrt <- glmLRT(fit, coef = 2)
```

We want to identify genes that show significantly different expression between the two strains. Look at the top 10 most significant results:
```{r}
topTags(lrt)
```

Get counts of significant genes (FDR-adjusted p-val < 0.01):
```{r}
tt.glm <- topTags(lrt, n = Inf)
nrow(tt.glm$table[tt.glm$table$FDR < 0.01, ])
```

Now grab the names of some highly significant genes:
```{r}
interestingSamples <- rownames(tt.glm$table[tt.glm$table$FDR < 1e-50, ])
cpm(dge.glm.tag.disp)[interestingSamples, ]
```

Characterize the number of genes altered in the strains:
```{r}
summary(de.glm <- decideTestsDGE(lrt, p = 0.05, adjust = "BH"))
```
There are `r length(which(de.glm == -1))` genes under-expressed in the DBA/2J mice, and `r length(which(de.glm == 1))` that are over-expressed. The remaining `r length(which(de.glm == 0))` show no significant differences in expression.

Observe the fold-changes:
```{r}
tags.glm <- rownames(dge.glm.tag.disp)[as.logical(de.glm)]
plotSmear(lrt, de.tags = tags.glm)
abline(h = c(-2, 2), col = "blue")
```

DESeq analysis
--------------
Now try the DE analysis using the DESeq package.

Create the data set:
```{r}
deSeqDat <- newCountDataSet(dat, group)
head(counts(deSeqDat))
```

Estimate size factors:
```{r}
deSeqDat <- estimateSizeFactors(deSeqDat)
sizeFactors(deSeqDat)
deSeqDat <- estimateDispersions(deSeqDat)
```

Plot dispersion vs.counts:
```{r}
plotDispEsts(deSeqDat)
```

Now fit the model:
```{r}
results <- nbinomTest(deSeqDat, levels(group)[1], levels(group)[2])
str(results)
plotMA(results)
```

Limma & voom analysis
---------------------
Use Voom and limma to perform DE analysis.

Prepare the data (figure out weights, etc.):
```{r}
norm.factor <- calcNormFactors(dat)
show(norm.factor)
dat.voomed <- voom(dat, design, plot = TRUE, lib.size = colSums(dat) * norm.factor)
dat.voomed
```

Now fit the model using our "voomed" data, and check out the top results:
```{r}
fit <- lmFit(dat.voomed, design)
fit <- eBayes(fit)
topTable(fit)
```