Seminar 4: Two group comparisons and data aggregation
========================================================
### Lauren Chong

Load in the data:
```{r}
library(lattice)
library(plyr)
library(xtable)
prDat <- read.table("GSE4051_data.tsv")
str(prDat, max.level = 0)

prDes <- readRDS("GSE4051_design.rds")
str(prDes)
```

### Two sample tests -- one gene
Randomly choose a gene, and extract its data:
```{r}
set.seed(540)
(myGene <- sample(1:nrow(prDat), size = 1))
pDat <- data.frame(prDes, gExp = unlist(prDat[myGene, ]))
str(pDat)
```

Explore the data a bit (try different summary fxns):
```{r}
aggregate(gExp ~ gType, pDat, FUN = mean)
ddply(pDat, ~ gType, summarize, expMean = mean(gExp))
ddply(pDat, ~ gType, function(x) return(c(expMean = mean(x$gExp))))
```

Do a sanity-check plot:
```{r}
stripplot(gExp ~ gType, pDat, pch = 19, col = "darkgrey", jitter = TRUE, ylab = "Expression", xlab = "Genotype")
```

Now try a t-test to see if the expression is different between genotypes:
```{r}
# Without using common variance assumption
t.test(gExp ~ gType, pDat)
# WITH common variance assumption
t.test(gExp ~ gType, pDat, var.equal = TRUE)
```

In both cases, the p-values indicate the genotype effect is not significant for this gene. Let's try other types of tests:
```{r}
wilcox.test(gExp ~ gType, pDat, exact = FALSE)
ks.test(x = pDat[pDat$gType == "wt", "gExp"],
        y = pDat[pDat$gType == "NrlKO", "gExp"],
        exact = FALSE)
```
Looks like the Wilcoxon and Kolmogorov-Smirnov tests agree that the effect is not significant.

### Two sample tests -- a handful of genes
Grab the data from 6 genes:
```{r}
keepGenes <- c("1431708_a_at", "1424336_at", "1454696_at",
               "1416119_at", "1432141_x_at", "1429226_at" )
miniDat <- subset(prDat, rownames(prDat) %in% keepGenes)
miniDat <- data.frame(gExp = as.vector(t(as.matrix(miniDat))),
                      gene = factor(rep(rownames(miniDat), each = ncol(miniDat)),
                                    levels = keepGenes))
miniDat <- suppressWarnings(data.frame(prDes, miniDat))
str(miniDat)
```

Plot to ensure we have some interesting and some boring genes:
```{r}
stripplot(gExp ~ gType | gene, miniDat,
          scales = list(y = list(relation = "free")),
          group = gType, auto.key = TRUE, jitter = TRUE)
```

Now do t-tests for these genes:
```{r}
ttRes <- ddply(miniDat, ~ gene, function(x) {
  res <- t.test(gExp ~ gType, x)
  return(c(pval = res$p.value))
})
```
```{r results="asis"}
foo <- xtable(ttRes)
print(foo, type = "html", include.rownames = FALSE)
```

### Concordance of two-group tests
Grab data for 100 genes at random:
```{r}
myGenes <- rownames(prDat)[sample(1:nrow(prDat), size = 100)]
myDat <- prDat[myGenes,]
myDat <- data.frame(
  gExp = as.vector(t(as.matrix(myDat))),
  gene = factor(rep(myGenes, each = ncol(myDat)), levels = myGenes)
)
myDat <- suppressWarnings(data.frame(prDes, myDat))
```

Now get p-values from t-test and Wilcoxon test for each gene:
```{r}
pValDat <- ddply(myDat, ~ gene, function(x) {
  ttRes <- t.test(gExp ~ gType, x)
  wRes <- wilcox.test(gExp ~ gType, x, exact = FALSE)
  return(c(ttPval = ttRes$p.value, wPval = wRes$p.value))
})
xyplot(ttPval ~ wPval,
       pValDat,
       panel = function(x,y,...) {
         panel.abline(a = 0, b = 1, col = "darkgrey")
         panel.xyplot(x,y,...)
       },
       ylab = "t-test p-value",
       xlab = "Wilcoxon p-value",
       pch = 19,
       col = "black"
)
```

Interesting! Looks like the p-values aren't so concordant. What about log-transformed?
```{r}
xyplot(ttPval ~ wPval,
       pValDat,
       panel = function(x,y,...) {
         panel.abline(a = 0, b = 1, col = "darkgrey")
         panel.xyplot(x,y,...)
       },
       ylab = "t-test p-value",
       xlab = "Wilcoxon p-value",
       pch = 19,
       col = "black",
       scales = list(x = list(log = TRUE), y = list(log = TRUE))
)
```
In particular, the less significant genes (higher p-values) are less concordant.